Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea

### 2023-02-13 analysis/paper/paper.qmd
edited up to results

abstract: |
  The onset of the Late Paleolithic period in Korea, represented by the appearance of stemmed points and blades, was a key event in the dispersal of modern humans in Northeast Asia. Previous studies have mainly focussed on possible origin locations of these new technologies. The specific cultural processes of the appearance stemmed points and blades has rarely been considered. We investigate the cultural processes by applying a cultural transmission framework to investigate the social contexts of the emergence of these new technologies. Our main question is: what was the dominant mode of cultural transmission during this time of technological innovation in the Korean Late Paleolithic? Following Bettinger and Eerkens [-@bettinger1999point], we evaluated models of guided variation and indirect bias using data from Korean assemblages containing stemmed points. To evaluate these models and understand the transmission processes, we computed correlation coefficients and coefficients of variation (CV). We found that information about the new technology was likely transmitted via guided variation with small impact of indirect bias. Some attributes including length and width were transmitted with less variation while other attributes appear to have more variation. Our results suggest that the dominant mode of cultural transmission for the earliest stemmed points was guided variation and then indirect bias was adapted to carry out the technological change. We assume that individuals or groups developed stemmed points by experimenting with existing blade technologies and then copied crucial part of a successful model to ensure the quality to optimize tool usage. As a result, the shape of stemmed points became more standardized among their social groups.

@@

Previous studies of the technological transitions during the Late Paleolithic in Korea have mainly focused on the possible origin locations of stemmed points, along with the issue of the timing and routes of modern human dispersal in eastern Asia. The debate about the origin of stemmed points can be summarized into two competing models: *in situ* evolution [@seong2009emergence] and heterogenic migration [@bae2010origin]. The *in situ* model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the Korean peninsula, as a form of convergent evolution [@seong2006structure; @seong2008tanged; @seong2009emergence]. To support his claim, @seong2009emergence examined the blade-to-flake ratios of stone artifact assemblages in South Korea. In his view, the blade industry represents a new technology while flakes indicate a continuously-used existing one. He argues that the increased ratio of blades in stone artifact assemblages during the Late Paleolithic shows an expansion of the new technology after its local invention. In addition, Seong claims that increased numbers of stemmed points over time, and standardization of their shape, supports the prominance of gradual, local, evolutionary processes in the emergence of new technologies.

In contrast, the migration model argues that the new blade industry, including stemmed points, and the earlier simple flake tool tradition, including large cores, polyhedrals, choppers and handaxes, came from different origins [@bae2010origin; @bae2012nature; @bae2013early]. While the *in situ* model claims that the heterogenic character is the result of indigenous development, the migration model proposes that it is the result of the continuous influx of modern human populations from both north and south. Specifically, the blade technology in claimed to have been introduced from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP, while the simple flake-tool tradition came from southern China [@bae2013early]. The migration of people from the southern route is supported by genetic studies of the Y chromosome, indicating that the O3-M122 M122 haplotype originated from southern East Asia and moved to northeastern Asia, including Korea, at 30-25 ka BP [@shi2005chromosome]. Bae et al. [-@bae2012nature] assume this southern migration could be related to paleoenvironmental fluctuations during the MIS 3 to 2 transition, which made the Yellow Sea/West Sea region open.

@@

Cultural transmission theory holds that information about how to behave is acquired through interaction with other individuals and the environment [@boyd1988culture; @richerson1992cultural]. Individuals learn by themselves (e.g. trial and error), or from each other by sharing information. Information can be modified (also known as 'biased') depending on an individual's transmission context and cultural repertoire. Modification of information can occur by recombination, loss, or partial alteration [@eerkens2005cultural; @o2017dual]. Transmission biases can be important loci of changes in material culture, and can be influenced by the social contexts of cultural transmission [@heyes1994social; @eerkens2007cultural; @lycett2015cultural; @creanza2017cultural; @kendal2018social]. These biases include guided variation (where individuals learn new behaviors through trial and error), content-based bias (where some aspect of the transmitted instructions, such as cultural preferences, makes them more likely to be adopted), frequency-based bias (where an individual is biased to choose particular behaviors based on their perceived frequency in the population, such as extremely popular or rare behaviors), and indirect bias (where a behavior is transmitted because of its association with other attributes, such as the prestige or skill of other individuals) [@boyd1988culture; @richerson1992cultural; @o2017dual].

@@

Following Bettinger and Eerkens' framework, Garvey [-@garvey2018current] uses simulation to explore the degree of standardization represented by coefficient of variation (CV) values of projectile points from the US Southwest and westernmost southern High Plains. Garvey measured CV values of weight, thickness, width and lengths on two types of projectile points, Washita and Fresno points, from the Henderson site. She then computed simulated CV values according to three scenarios of different levels of transmission fidelity (i.e. CV = 10%, 5%, and 3%). The observed metrics of the archaeological projectile points are closest to the metrics of the simulated assemblage with 3% CV, which represents "extremely high-fidelity copying". A CV of 3% is the Weber fraction, or threshold of human visual perception, with variation below this value being too subtle for people to notice [@eerkens2000practice]. Garvey's work shows how the cultural transmission of tool-making behaviours can be measured, simulated, and used to interpret the archaeological record.

@@

Inspired by Bettinger and Eerkens' approach, we use two contrasting transmission modes, guided variation and indirect bias, to investigate the spread of stemmed point technology during the Korean Late Paleolithic period. We propose a spectrum on which we can locate foraging groups and how they started to make stemmed points, depending on their degree of social isolation or social connectedness. On one end of the spectrum we have socially isolated groups who made stemmed points entirely stimulated by guided variation; and on the other end of the spectrum we have socially connected groups whose knowledge of stemmed points derived from indirect bias on the transmission process.

To contextualize this further, on one end of our spectrum we have socially isolated groups who stayed in physically remote places from other groups, or had unfamiliar neighbors with limited contact between groups. Our assumption for this social context is that individuals or groups acquired the technology of stemmed points by modifying existing flake tool forms through trial-and-error processes to solve problems relating to resource procurement (cf. Seong's *in situ* model). This trial and error behavior is expected to have left a distinctive signature on metric variables of the stemmed points. If trial and error was the dominant bias in the transmission of knowledge about how to make stemmed points, we expect that the morphological attributes on a stemmed point will be poorly correlated and show high variation within assemblages and between sites.

On the other end of our spectrum, we have socially connected groups that occupied places close to other groups or had regular contact with other groups. Relatively high degrees of social connectivity provide frequent opportunities for observing others and acquiring information. Individuals or groups learned the technology of stemmed points by copying a model from another individual or group. In this scenario, the model, or ideal, stemmed point is likely to have been a highly successful and frequently chosen one, and learners copy all information about the point design as a package (cf. Bae's migration model). In this social context, indirect bias is the dominant influence on transmission of lithic technology. If indirect bias was prominent in the Korean Paleolithic, we expect that stemmed points should be more standardized, and attributes on the stemmed points will be more correlated, and assemblages from multiple sites will show less variance between and within them.

@@

To analyze morphological change over time, we used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013]. These chronologies were built on radiocarbon ages, and for sites without any, the presence/absence of blades, stemmed point blanks, and toolkit composition. Based on the existence of blade and micro blade and types of stemmed point blanks, we divided the Korean Late Paleolithic assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blade or flake and the existence of blade in assemblages, and 3) stemmed points made out of blade and the existence of micro blade in assemblages (@tbl-korean-chronology). Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon ages due to various research limitations. Inferring ages for these sites by analogy to the technological sequences at sites with dates is important for making maximal use of the available archaeological data to increase our sample size. We arranged our `r total_number_of_assemblages` assemblages containing stemmed points in into these three phases to facilitate observations of change over time.

```{r}
#| label: tbl-korean-chronology
#| tbl-cap: Korean Late Paleolithic Chronology edited based on Seong [-@seong2015diversity] and Park [-@Park_2013]

@@

Panel A of @fig-raw-materials-by-phase summarizes the amount of stemmed points in each chronological phase, and the distribution of raw materials across the phases. We exclude phase 1 from our analyses below because of the small sample size to compute correlation and CV. Shale is the dominant raw material in both phase 2 and 3. In phase 3 there is an increase in the proportions of porphyry, driven largely by finds from Yongsangdong, and other raw materials. Panel B in @fig-raw-materials-by-phase shows the count of stemmed points by raw material type at assemblages with more than five stemmed points. Much of the raw material diversity comes from isolated finds, with those assemblages with more than five points showing high homogeneity in raw materials.

@@

Similar to the metric attributes from previous studies of cultural transmission and projectile points, i.e. maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we used morphological attributes on the stemmed points to examine variations in shape and the relationship between each attribute (@fig-labelled-schematic). We obtained our morphological attribute data from landmark analysis of digitized images of stemmed points. Compared to traditional caliper measurements, morphometric data yields more easily interpretable numerical and visual outcomes [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life]. We did not include weight and thickness as these measurements were not available to us. For the landmark analysis, we put a total of 11 landmarks on the outline of each stemmed point and computed distances between pairs of landmark coordinates to derive attributes for our statistical analyses. The landmarks we recorded are described in @tbl-landmark-abbreviations and shown in @fig-labelled-schematic. We used the point tool in ImageJ [@schneider2012nih] to capture the landmarks from images of the artifacts, and export them as XY coordinate data for current and further analyses.

@@

We chose Coefficients of Variation (CV) to measure the variation among the measurements of the stemmed points. The CV is the ratio (%) of the standard deviation to the mean. It has been used as a robust statistical method to calculate the degree of standardization, precision, equality, homogeneity, etc. in various disciplines [@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence]. In archaeological studies, CV has been applied to measure the variation between artifacts and to test hypotheses about cultural evolutionary processes, including distinguishing between different types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current]. Based on the Weber Fraction for visual measurement of line length, Eerkens and Bettinger [-@eerkens2001techniques] claim that over 57.7% of the CV is the result of random production and below 1.7% is the byproduct of using a scale or template.

Sample size can impact the accuracy of CV estimates [@toebe2018sample; @kelley2007sample; @vanpool2011quantitative]. In previous archaeological studies, sample sizes for CV values has varied from five to more than a thousand artifacts [@bettinger1999point; @garvey2018current; @costin1995standardization; @wierer2013variability; @kvamme1996alternative; @rivals2009new]. A variety of methods are available for improving the reliability of CV measurements when using small samples. For example, @vanpool2011quantitative proposed a "corrected CV", for a smaller sample (i.e. n\<25). Statistical research has resulted in several methods for computing confidence intervals on CVs to show uncertainty [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison]. Although we have a total of `r total_number_of_sp` artifacts, most of our analyses here involves comparisons of smaller subsets of this sample. For example to explore temporal and regional patterns we compared samples of well under 25 pieces. Given these relatively small sample sizes in our study, we used the corrected CV formula, which we have implemented here in a function in the R programming language for others to use. We also computed confidence intervals using method proposed by @sharma1994asymptotic to increase the credibility of the comparisons and be transparent about the precision and accuracy of our results [@kelley2007sample; @albatineh2014confidence].

@eerkens2001techniques presented some average CV values (%) of artifacts made from various materials (See Table 1 from @eerkens2001techniques). 
Among those various material artifacts, the most comparable artifacts to the stemmed points are Great Basin projectiles points, whose CV value ranged from 6-55, with an average of 22. Also, projectile points from the US Southwest have CV values ranged from 11 to 33 [@garvey2018current]. Unlike correlation, @bettinger1999point had vaguely mentioned about the threshold for CV values. Therefore, here we sat `r threshold_value` as a rough threshold to evaluate the transmission biases referring to the example of Great Basin projectile points. We calculated CV values of Rosegate points from Monitor Valley, which was speculated as the byproducts of indirect bias (raw data from table 5 of @bettinger1999point). With the exception of weight, the CV range for metric attributes on the points is 17.01 to 24.59. We chose higher side on the CV range of Rosegate points as the threshold of our research based on following reasons. 1) Variation is generated by small errors that are transmitted between individuals and the errors gets bigger through generations [@eerkens2005cultural]. The duration of transmission process for stemmed points is much longer (approx.30,000yrs) than the case of Rosegate (approx.500yrs in Fremont region) [@bischoff2021rosegate]. 2) The errors are also highly associated with raw materials. Some raw materials, such as clay, are easier to control while, less predictable and controllable materials such as stone are likely to have inflated CV values [@eerkens2001techniques]. We assume that Rosegate projectile points were made from more finer grained raw materials such as flint compared to raw materials for stemmed points such as shale, rhyolite, porphyry, etc. (@fig-raw-materials-by-phase). 3) Compared to normal CV, corrected CV has slightly higher values [@vanpool2011quantitative]. Based on the threshold, we premised that CV values that are lower than `r threshold_value` represent guided variation while CV values over `r threshold_value` reflect indirect bias.

@@

###### facet plot
cv_plot_raw_materials_label %>%
  ggplot(aes(x = variable, y = cv_by_materials)) +
  geom_point() +
  geom_errorbar(aes(ymin = cv_low_sharma,
                    ymax = cv_high_sharma), width = .2) +
    # draw a line to show our threshold value
  geom_hline(yintercept = 25, colour = "red") +
  xlab("Raw Materials") +
  ylab("Coefficient of Variation on Attributes") +
  facet_wrap( ~ label, scales = "free_y") +
  coord_cartesian(ylim = c(0,250)) +
  theme_bw() 
```
# Discussion

@@

```
In exploring geographical variation we found no clear pattern variation in both correlation coefficients (@fig-plot-cor-four-assemblage) and CV values (@fig-plot-cv-four-assemblage) between assemblages with more than five stemmed points. We found that each assemblage has different distribution patterns of correlation coefficients and CV values. SYG1_2 and SYG1_3 are from the same site but different chronological periods and correlations decreases over time. It is likely that correlation or overall shape can be more depending on individual knappers. We also found that some attributes such as maximum width and maximum length have stronger correlation while the others including stemmed length and body length have lower or even negative correlation. We assume that certain attributes were more carefully transmitted than others. We observed that CV values of body length, maximum length, mid width, and tang width are generally lower than tang length, stem length, and stem width, similar to what is evident in @fig-plot-cv-all-attributes. The attributes with lower CV values could be transmitted in a social context dominated by indirect bias, while the higher CV attributes could be influenced more by specific manufacturing and maintenance situations. We assume that the lower CV values of the four attributes could imply that those four attributes are closely associated with the projectile function of the tool so that the knappers wanted to standardize the shape. For example, body length is related to penetration, durability, hardness, and rejuvenation of a projectile point and tang width is connected to wound damage, penetration, durability and hardness [@odell1986experiments; @shea2001experimental; @cheshier2006projectile; @yaroshevich2016projectile; @bebber2017developing; @wood2018wound]. This can be contrasted with tang length, stem length, and stem width, which are away from the impact end of the artifact. These were allowed to vary more freely between points, perhaps to be able to accommodate shafts of different types of wood with varying properties of strength, flexibility, and weight.

>>>>

abstract: |
  The onset of the Late Paleolithic period in Korea, represented by the appearance of stemmed points and blades, was a key event in the dispersal of modern humans in Northeast Asia. Previous studies have mainly focussed on possible origin locations of these new technologies. The specific cultural processes of the appearance stemmed points and blades has rarely been considered. We investigate the cultural processes by applying a cultural transmission framework to investigate the social contexts of the emergence of these new technologies. Our main question is: what was the dominant mode of cultural transmission during this time of technological innovation in the Korean Late Paleolithic? Following Bettinger and Eerkens [-@bettinger1999point], we evaluated models of guided variation and indirect bias using data from Korean assemblages containing stemmed points. To evaluate these models and understand the transmission processes, we computed correlation coefficients and coefficients of variation (CV). We found that information about the new technology was likely transmitted via guided variation with small impact of indirect bias. Some attributes including length and width were transmitted with less variation while other attributes appear to have more variation. Our results suggest that the dominant mode of cultural transmission for the earliest stemmed points was guided variation <!-- and then indirect bias was adapted to carry out the technological change BM: this is unclear -->. We assume that individuals or groups developed stemmed points by experimenting with existing blade technologies and then copied crucial parts of a successful model to ensure the quality to optimize tool usage. As a result, the shape of stemmed points became more standardized among their social groups.

@@

Previous studies of the Late Paleolithic technological transitions in Korea have mainly focused on the possible origin locations of stemmed points, as part of the discussion of the timing and routes of modern human dispersal in eastern Asia. The debate about the origin of stemmed points can be summarized into two competing models: *in situ* evolution [@seong2009emergence] and heterogenic migration [@bae2010origin]. The *in situ* model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the Korean peninsula, as a form of convergent evolution [@seong2006structure; @seong2008tanged; @seong2009emergence]. To support his claim, @seong2009emergence examined the blade-to-flake ratios of stone artifact assemblages in South Korea. In his view, the blade industry represents a new technology while flakes indicate a continuously-used existing one. He argues that the increased ratio of blades in stone artifact assemblages during the Late Paleolithic shows an expansion of the new technology after its local invention. In addition, Seong claims that increased numbers of stemmed points over time, and standardization of their shape, supports the prominance of gradual, local, evolutionary processes in the emergence of new technologies.

In contrast, the migration model argues that the new blade industry, including stemmed points, and the earlier simple flake tool tradition, including large cores, polyhedrals, choppers and handaxes, came from different origins outside of the Korean peninsula [@bae2010origin; @bae2012nature; @bae2013early]. While the *in situ* model claims that the heterogenic character is the result of indigenous development, the migration model proposes that it is the result of the continuous influx of modern human populations from both north and south. Specifically, the blade technology is claimed to have been introduced from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP, while the simple flake-tool tradition came from southern China [@bae2013early]. The migration of people from the southern route is supported by genetic studies of the Y chromosome, indicating that the O3-M122 M122 haplotype originated from southern East Asia and moved to northeastern Asia, including Korea, at 30-25 ka BP [@shi2005chromosome]. Bae et al. [-@bae2012nature] assume this southern migration could be related to paleoenvironmental fluctuations during the MIS 3 to 2 transition, which made the Yellow Sea/West Sea region open.

@@

Cultural transmission theory holds that information about social and technological behaviours is acquired through interaction with other individuals and the environment [@boyd1988culture; @richerson1992cultural]. Individuals learn by themselves (e.g. trial and error), or from each other by sharing information. Information can be modified (also known as 'biased') depending on an individual's transmission context and cultural repertoire. Modification of information can occur by recombination, loss, or partial alteration [@eerkens2005cultural; @o2017dual]. Transmission biases can be important loci of changes in material culture, and can be influenced by the social contexts of cultural transmission [@heyes1994social; @eerkens2007cultural; @lycett2015cultural; @creanza2017cultural; @kendal2018social]. These biases include guided variation (where individuals learn new behaviors through trial and error), content-based bias (where some aspect of the transmitted instructions, such as cultural preferences, makes them more likely to be adopted), frequency-based bias (where an individual is biased to choose particular behaviors based on their perceived frequency in the population, such as extremely popular or rare behaviors), and indirect bias (where a behavior is transmitted because of its association with other attributes, such as the prestige or skill of other individuals) [@boyd1988culture; @richerson1992cultural; @o2017dual].

@@

Following Bettinger and Eerkens' framework, Garvey [-@garvey2018current] uses simulation to explore the degree of standardization represented by coefficient of variation (CV) values of projectile points from the US Southwest and westernmost southern High Plains. Garvey measured CV values for weight, thickness, width and lengths on two types of projectile points, Washita and Fresno points, from the Henderson site. She then computed simulated CV values according to three scenarios of different levels of transmission fidelity (i.e. CV = 10%, 5%, and 3%). The observed metrics of the archaeological projectile points are closest to the metrics of the simulated assemblage with 3% CV, which represents "extremely high-fidelity copying". A CV of 3% is the Weber fraction, or threshold of human visual perception, with variation below this value being too subtle for people to notice [@eerkens2000practice]. Garvey's work shows how the cultural transmission of tool-making behaviours can be measured, simulated, and used to interpret the archaeological record.

@@

Inspired by Bettinger and Eerkens' approach, we use two contrasting transmission modes, guided variation and indirect bias, to investigate the spread of stemmed point technology during the Korean Late Paleolithic period. We propose a spectrum on which we can locate foraging groups and how they started to make stemmed points, depending on their degree of social isolation or social connectedness. On one end of the spectrum we have socially isolated groups who made stemmed points through guided variation; and on the other end of the spectrum we have socially connected groups whose knowledge of stemmed points derived from transmission processes dominated by indirect bias.

To contextualize this further, on one end of our spectrum we have socially isolated groups who stayed in physically remote places from other groups, or had unfamiliar neighbors with limited contact between groups. Our assumption for this social context is that individuals or groups acquired the technology of stemmed points by modifying existing flake tool forms through trial-and-error processes to solve problems relating to resource procurement (cf. Seong's *in situ* model). If this social context was prevalent, we predict this trial and error behavior could have left a distinctive signature on the metric variables of the stemmed points. If trial and error was the dominant bias in the transmission of knowledge about how to make stemmed points, we expect that the morphological attributes on a stemmed point will be poorly correlated and show high variation within assemblages and between sites.

On the other end of our spectrum, we have socially connected groups that occupied places close to other groups or had regular contact with other groups. Relatively high degrees of social connectivity provide frequent opportunities for observing others and acquiring information. Individuals or groups learned the technology of stemmed points by copying a model from another individual or group. In this scenario, the model, or ideal, stemmed point is likely to have been a highly successful and frequently chosen one, and learners copy all information about the point design as a package. This implies a small number of locations, or a single point (perhaps from a source external to the Korean peninsula, cf. Bae's migration model), where stemmed points first appeared and then spread from. In this social context, indirect bias is the dominant influence on transmission of lithic technology. If indirect bias was prominent in the Korean Paleolithic, we expect that stemmed points should be more standardized, and attributes on the stemmed points will be more correlated, and assemblages from multiple sites will show less variance between and within them.

@@

To analyze morphological change over time, we used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013]. These chronologies were built on radiocarbon ages, and for sites without any, the presence/absence of blades, stemmed point blanks, and toolkit composition were used to assign assemblages to a phase. Following these previous chronological schemes, we divided the Korean Late Paleolithic assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blades or flakes and the existence of blade in assemblages, and 3) stemmed points made out of blades and the existence of micro blades in assemblages (@tbl-korean-chronology). Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon ages due to research limitations. Inferring ages for these sites by analogy to the technological sequences at sites with dates is necessary for making maximal use of the available archaeological data to increase our sample size. We arranged our `r total_number_of_assemblages` assemblages containing stemmed points in into these three phases to facilitate observations of change over time.

```{r}
#| label: tbl-korean-chronology
#| tbl-cap: Korean Late Paleolithic Chronology edited based on Seong [-@seong2015diversity] and Park [-@Park_2013]. SP = stemmed point

@@

Panel A of @fig-raw-materials-by-phase summarizes the amount of stemmed points in each chronological phase, and the distribution of raw materials across the phases. We exclude phase 1 from our analyses below because of the small sample size to compute correlations and CVs. Shale is the dominant raw material in both phase 2 and 3. In phase 3 there is an increase in the proportions of porphyry, driven largely by finds from Yongsangdong, and other raw materials. Panel B in @fig-raw-materials-by-phase shows the count of stemmed points by raw material type at assemblages with more than five stemmed points. Much of the raw material diversity comes from isolated finds, with those assemblages with more than five points showing high homogeneity in raw materials.

@@

Following the metric attributes used in previous studies of cultural transmission and projectile points, i.e. maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we used morphological attributes on the stemmed points to examine variations in shape and the relationship between each attribute (@fig-labelled-schematic). We obtained our morphological attribute data from landmark analysis of digitized images of stemmed points. Compared to traditional caliper measurements, morphometric data yields more easily interpretable numerical and visual outcomes than traditional caliper measurements [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life]. We did not include weight and thickness as these measurements were not available to us. For the landmark analysis, we put a total of 11 landmarks on the outline of each stemmed point and computed distances between pairs of landmark coordinates to derive attributes for our statistical analyses. The landmarks we recorded are described in @tbl-landmark-abbreviations and shown in @fig-labelled-schematic. We used the point tool in ImageJ [@schneider2012nih] to capture the landmarks from images of the artifacts, and export them as XY coordinate data for further analyses.

@@

We chose Coefficients of Variation (CV) to measure the variation among the measurements of the stemmed points. The CV is the ratio (usually expressed as a percentage) of the standard deviation to the mean. It has been used as a robust statistical method to calculate the degree of standardization, precision, equality, homogeneity, etc. in various disciplines [@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence; @wang2020standardization]. In archaeological studies, CV has been applied to measure the variation between artifacts and to test hypotheses about cultural evolutionary processes, including distinguishing between different types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current]. As a guide to interpreting CV values, Eerkens and Bettinger [-@eerkens2001techniques] claim that over 57.7% of the CV is the result of random production and below 1.7% is the by-product of using a scale or template. Among the various artifacts for which \@eerkens2001techniques summarised CV values, the most comparable to stemmed points are Great Basin projectiles points, whose CV values range from 6-55%, with an average of 22%. Similarly, projectile points from the US Southwest have CV values ranging from 11% to 33% [@garvey2018current].

Sample size can impact the accuracy of CV estimates [@toebe2018sample; @kelley2007sample; @vanpool2011quantitative]. In previous archaeological studies, sample sizes for CV values has varied from five to more than a thousand artifacts [@bettinger1999point; @garvey2018current; @costin1995standardization; @wierer2013variability; @kvamme1996alternative; @rivals2009new]. A variety of methods are available for improving the reliability of CV measurements when using small samples. For example, @vanpool2011quantitative proposed a "corrected CV", for a smaller samples (i.e. n\<25). Statistical research has resulted in several methods for computing confidence intervals on CVs to show uncertainty [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison]. Although we have a total of `r total_number_of_sp` artifacts, most of our analyses here involves comparisons of smaller subsets of this sample. For example to explore temporal and regional patterns we compared samples of well under 25 pieces. Given these relatively small sample sizes in our study, we used the corrected CV formula, which we have implemented here in a function in the R programming language for others to use. We also computed confidence intervals using method proposed by @sharma1994asymptotic to increase the credibility of the comparisons and be transparent about the precision and accuracy of our results [@kelley2007sample; @albatineh2014confidence].

One limitation of previous work is the absence of a clear threshold value for interpreting CV values in terms of different transmission biases. @bettinger1999point did not identify a threshold for CV values. Here we propose `r threshold_value` as an approximate threshold to distinguish between the transmission biases. We used data in @bettinger1999point to calculate CV values of Rosegate points from Monitor Valley, California, which was speculated by Bettinger and Eerkens to be the byproducts of indirect bias. With the exception of weight, the CVs for metric attributes on the Rosegate points is 17% to 24%. We believe higher side of the CV range of Rosegate points is appropriate as a threshold value for our research based on following reasons. First, variation is generated by small errors that are transmitted between individuals and the errors gets bigger through generations [@eerkens2005cultural]. The duration of transmission process for stemmed points is much longer (in the order of thousands of years) than the case of Rosegate (in the order of hundreds of years in Fremont region) [@bischoff2021rosegate]. Second, the errors are likely to vary by raw materials. Some raw materials, such as clay, are easier to control variation in while less controllable materials such as stone are likely to have higher CV values because of the relatively unpredictable nature of flaking [@eerkens2001techniques]. We assume that Rosegate projectile points were made from more finer grained raw materials such as flint compared to the raw materials for Korean stemmed points, which include shale, rhyolite, porphyry, etc. (@fig-raw-materials-by-phase). Third, the corrected CV calculation for small sample sizes that we are using tends to result in slightly higher values than compared to standard CV values [@vanpool2011quantitative]. Given the tentative identification of this threshold value, we premised that CV values that are lower than `r threshold_value` likely represent guided variation while CV values over `r threshold_value` likley reflect indirect bias.

@@

###### facet plot
cv_plot_raw_materials_label %>%
  ggplot(aes(x = variable, y = cv_by_materials)) +
  geom_point() +
  geom_errorbar(aes(ymin = cv_low_sharma,
                    ymax = cv_high_sharma), width = .2) +
    # draw a line to show our threshold value
  geom_hline(yintercept = 25, colour = "red") +
  xlab("Raw Materials") +
  ylab("Coefficient of Variation on Attributes") +
  facet_wrap( ~ label, scales = "free_y") +
  coord_cartesian(ylim = c(0,250)) +
  theme_bw() 
```

# Discussion

@@

In exploring geographical variation we found no clear pattern variation in both correlation coefficients (@fig-plot-cor-four-assemblage) and CV values (@fig-plot-cv-four-assemblage) between assemblages with more than five stemmed points. We found that each assemblage has different distribution patterns of correlation coefficients and CV values. SYG1_2 and SYG1_3 are from the same site but different chronological periods and correlations decreases over time. It is likely that correlation or overall shape can be more depending on individual knappers. We also found that some attributes such as maximum width and maximum length have stronger correlation while the others including stemmed length and body length have lower or even negative correlation. We assume that certain attributes were more carefully transmitted than others. We observed that CV values of body length, maximum length, mid width, and tang width are generally lower than tang length, stem length, and stem width, similar to what is evident in @fig-plot-cv-all-attributes. The attributes with lower CV values could be transmitted in a social context dominated by indirect bias, while the higher CV attributes could be influenced more by specific manufacturing and maintenance situations. We assume that the lower CV values of the four attributes could imply that those four attributes are closely associated with the projectile function of the tool so that the knappers wanted to standardize the shape. For example, body length is related to penetration, durability, hardness, and rejuvenation of a projectile point and tang width is connected to wound damage, penetration, durability and hardness [@odell1986experiments; @shea2001experimental; @cheshier2006projectile; @yaroshevich2016projectile; @bebber2017developing; @wood2018wound]. This can be contrasted with tang length, stem length, and stem width, which are away from the impact end of the artifact. These were allowed to vary more freely between points, perhaps to be able to accommodate shafts of different types of wood with varying properties of strength, flexibility, and weight.
